Representations of rational Cherednik algebras of $\Sigma_n$ in positive characteristic

### 2015-05-03 formalwriteup.tex
paper draft

\section{Introduction and Definitions}

@@

%motivation, history, etc.

Given a vector space $\h$, an element $s \in \GL(\h)$ is an {\it reflection} if it has finite order and $\rank(1-s)=1$. A finite subgroup of $\GL(\h)$ that is generated by reflections is a {\it reflection group}. In particular the symmetric group $\Sigma_n$ in $n$ variables is a reflection group (the reflections are the transpositions).

Given a reflection group $G \subset \GL(\h)$ and a vector space $\h$ over a field $k$, we let $\mathcal{S}$ be the set of reflections in $G$. For each $s \in \mathcal{S}$ we assign a vector $\alpha_s \in \h^*$ spanning the image of $1-s$, and choose $\alpha_s^\vee \in \h$ so that $(1-s)x=\langle \alpha_s^\vee,x\rangle \alpha_s$ for all $x \in \h^*$, where $\langle \cdot,\cdot\rangle$ indicates the pairing between $\h$ and $\h^*$. We choose $\hbar \in k$ a number and $c: \mathcal{S} \to k$ a function so that $c(s)=c(s')$ whenever $s,s'$ are conjugate. Let $\overline{c}$ be the function defined by $\overline{c}(s)=c(s^{-1})$.

Let $T(\h \oplus \h^*)$ be the tensor algebra of $\h \oplus \h^*$. Then we define the {\it rational Cherednik algebra $\HH_{\hbar,c}(G,\h)$} as the quotient of $k[G] \ltimes T(\h \oplus \h^*)$ by the following relations for all $x,x' \in \h^*, y,y' \in \h$:
\begin{align*}
[x,x']=0, \quad [y,y' = 0], \quad [y,x] = \hbar\langle y,x\rangle - \sum_{s \in \mathcal{S}} c(s)\langle y,\alpha_s\rangle\langle \alpha_s^\vee,x\rangle s.
\end{align*}
We can give $\HH_{\hbar,c}(G,\h)$ a $\mathbb{Z}$-grading by setting $\deg x=1$ for $x \in \h^*$, $\deg y = -1$ for $y \in \h$, and $\deg g=0$ for $g \in k[G]$. We get the PBW-type decomposition $\HH_{\hbar,c}(G,\h)=\Sym(\h) \otimes_k k[G] \otimes_k \Sym(\h^*)$ (\cite{EM}, section 3.2).

In general, for any $\alpha \ne 0$, $\HH_{\hbar,c}(G,\h)\simeq \HH_{\alpha\hbar,\alpha c}(G,\h)$. Then we can assume $\hbar=0$ or $\hbar = 1$.

Let $\tau$ be a representation of $G$. The {\it Verma module} $M_{\hbar,c}(G,\h,\tau)$ is defined as $\HH_{\hbar,c}(G,\h) \otimes_{k[G] \ltimes \Sym(\h)} \tau$. Using the PBW decomposition of the Cherednik algebra, we see that $M_{\hbar,c}(G,\h,\tau)=\Sym(\h^*) \otimes_k \tau$ as a $k$-vector space; we can give this a $\mathbb{Z}$-grading in an obvious way.

As described in section 2.5 of \cite{BC1}, $M_{\hbar,c}(G,\h,\tau)$ has a unique maximal graded proper submodule $J_{\hbar,c}(G,\h,\tau)$ which can be realized as the kernel of the contravariant form $\beta_c: M_{\hbar,c}(G,\h,\tau) \times M_{\hbar,\overline{c}}(G,\h^*,\tau^*) \to k$; $\beta_c$ can be characterized by the property that for all $x \in \h^*, y \in \h, f \in M_{\hbar,c}(G,\h,\tau), g \in  M_{\hbar,\overline{c}}(G,\h^*,\tau^*), v \in \tau, w \in \tau^*$:
\begin{align*}
\beta_c(xf,g)=\beta_c(f,xg), \quad \beta_c(f,yg) = \beta_c(yf,g), \quad \beta_c(v,w) = \langle v,w\rangle.
\end{align*}
The quotient $L_{\hbar,c}(G,\h,\tau) = M_{\hbar,c}(G,\h,\tau)/J_{\hbar,c}(G,\h,\tau)$ is a finite-dimensional irreducible $\mathbb{Z}$-graded representation of $\HH_{\hbar,c}(G,\h)$.

To understand the action of $\HH_{\hbar,c}(G,\h)$ on $M_{\hbar,c}(G,\h,\tau)$, we can use the PBW decompositions. The action of $\Sym(\h^*)$ on $M_{\hbar,c}(G,\h,\tau)=\Sym(h^*)\otimes_k \tau$ is by left multiplication; $k[G]$ acts by the diagonal action, and $\Sym(\h)$ acts via {\it Dunkl operators}. For $y \in \h$, the Dunkl operator $D_y$ acts on $M_{\hbar,c}(G,\h,\tau)$ by:
\begin{align*}
D_y(f \otimes v) = \hbar \partial_y f \otimes v - \sum_{s \in \mathcal{S}} c(s) \frac{ ( y, \alpha_s )}{\alpha_s} (1-s). f \otimes s.v.
\end{align*}
Throughout the paper we let $G=\Sigma_n$ and $\tau$ the trivial representation. There is only one conjugacy class of reflections in $\Sigma_n$, so $c$ is an element of $k$ for our purposes. We call $c$ {\it generic} if we do not specify its value. We will be concerned with the case $\hbar=1$ and $c$ generic in this paper. (Note that in particular this case means that $\overline{c}=c$, since there is only one conjugacy class of reflections.) The characteristic of the field $k$ is $p>0$. We let $V$ be the vector space spanned by $y_0,\dots,y_{n-1}$ and let $\h$ be the subspace spanned by $y_i-y_j$ for $i \ne j$; $\Sigma_n$ acts by permuting indices. Then if $x_0,\dots,x_{n-1}$ is the dual basis for $V^*$, we see that $\h^*$ is the span of $x_0,\dots,x_{n-1}$ under the relation $x_0+\dots+x_{n-1}=0$; alternatively we can consider $\h^*$ as the span of $x_0,\dots,x_{n-2}$ with $x_{n-1}$ defined as $-x_0-\dots-x_{n-2}$. For a transposition $s_{ij} \in \Sigma_n$ with $i<j$, we let the corresponding vector $\alpha_{s_{ij}} \in \h^*$ be $x_i-x_j$.

In this case, since $\tau$ is trivial, $M_{\hbar,c}(G,\h,\tau)$ is a polynomial ring $k[x_0,\dots,x_{n-2}]$; we call this polynomial ring $A$. Similarly we refer to $J_{\hbar,c}(G,\h,\tau)$ as $J$, which is an ideal in $A$, and $A/J$ is the irreducible representation of the Cherednik algebra we desire to find. Because of the definition of the contravariant form $\beta_c$, showing that an element $f$ of $A$ is in the kernel of $\beta_c$ is equivalent to showing that the Dunkl operators corresponding to the basis elements of $\h$ annihilate $f$. If $D_i$ is the Dunkl operator corresponding to $y_i$, the Dunkl operators for the elements of $\h$ are spanned by $D_i-D_j$ with $i \ne j$.

We assume $p>2$ since the case $p=2$ is fully characterized in \cite{L}.

@@

In this paper we consider the case where the characteristic $p$ of $k$ divides $n$. This is related to the case in characteristic $0$ where $c$ takes the specific value $p/n$, as described in \cite{CE}. In that case the generators of the ideal $J$ were the residues at infinity of $\frac{1}{z-x_a} \prod_{i=0}^{n-1} (z-x_i)^c$ for each $a$. To get a similar result in positive characteristic, we must consider formal power series in $z$ with coefficients from $A$. The formal power series would be $\frac{1}{z^{p+1}}\frac{1}{1-x_az}\prod_{j=0}^{n-1} (1-x_jz)^c$, with the corresponding generator as the coefficient of $1/z$. We simplify and truncate the formal power series so we can define it in positive characteristic.

@@

@@

\begin{proposition}\label{prop:linind} The $f_a$ for $a=0,\dots,n-2$ are linearly independent homogeneous polynomials of degree $p$.%, with $f_{n-1}=-\sum_{a=0}^{n-2} f_a$. 
\end{proposition}

@@

%We also note that $\frac{\partial g}{\partial z} = g\left(\sum_a\frac{-x_az}{1-x_az}\right)$. Then

%\begin{align*}
%\sum_{a=0}^{n-1} F_a&=\left(\sum_a \frac{1}{1-x_az}\right)\left(\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k\right)\\
%&=\left(\sum_a \frac{1}{1-x_az}\right)\left(\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k\right)-n\left(\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k\right)\\
%&=\left(\sum_a\frac{x_az-1}{1-x_az}+ \frac{1}{1-x_az}\right)\left(\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k\right)\\
%&=\left(\sum_a\frac{-x_az}{1-x_az}\right)\left(\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k\right)\\
%&=\frac{z}{cg}\frac{\partial g}{\partial z}\left(\sum_{k=0}^{p-1} c\binom{c}{k} (g-1)^k\right)\\
%&=\frac{z}{c}\frac{\partial g}{\partial z}\left(\sum_{k=1}^{p-1} k\binom{c}{k} (g-1)^{k-1}\right)\\
%&=\frac{z}{c}\frac{\partial }{\partial z}\left(\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k\right)
%\end{align*}

%This equality only holds up to the $z^p$ coefficient, since we implicitly add a multiple of $(g-1)^{p-1}$ in the second-to-last step.

%Then the coefficient of $z^p$ in this sum is $1/c$ times the coefficient of $z^{p-1}$ in $\frac{\partial }{\partial z}\left(\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k\right)$, which is $p$ times the coefficient of $z^p$ in $\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k$, which must be $0$ since we are in characteristic $p$. The coefficient of $z^p$ in this sum is also $\sum_{a=0}^{n-1} f_a$, so we have $\sum_{a=0}^{n-1} f_a=0$, and $f_{n-1}=-\sum_{a=0}^{n-2} f_a$.

@@

We write $x$ for the vector $\langle x_0,\dots, x_{n-2} \rangle$, where the $x_i$ are taken from the rational function field in $c$ over $k$. Then we can consider $f_a$ as a function on these vectors $x$ for all $a$. For any rational function $u( c)$, we let $u(c )x=\langle u(c )x_0,\dots, u(c )x_{n-2}\rangle$.

@@

\begin{theorem}\label{thm:main} The $f_a$ generate the ideal $J$; $A/J$ has Hilbert series $\left(\frac{1-t^p}{1-t}\right)^{n-1}$. 
\end{theorem}

\begin{proof} By Propositions~\ref{prop:linind}, \ref{prop:ci}, $A/I$ is a complete intersection with $n-1$ generators of degree $p$. It then must have Hilbert series $h_{A/I}(t)=\left(\frac{1-t^p}{1-t}\right)^{n-1}$. By Proposition~\ref{prop:ann}, the generators of $I$ are annihilated by the Dunkl operators, so $I \subseteq J$.

@@

%something about how it's a model problem

>>>>

\section{Introduction}

@@

This problem is an interesting test case for moving results from characteristic $0$ to positive characteristic. Cherednik algebras, originally double affine Hecke algebras, were introduced by Cherednik in 1993; for an introduction to these algebras we refer to \cite{EM}. The representation theory of these algebras has been studied more extensively in characteristic $0$ than in positive characteristic, including the case of Cherednik algebras of type A (associated to the symmetric group $\Sigma_n$). In general new techniques are required to study positive characteristc as opposed to characteristic 0. In this paper we directly connect the characteristc 0 case to the positive characteristic case; instead of taking complex residues, we consider the coefficients of formal power series in positive characteristic. While we only consider the case of the symmetric group $\Sigma_n$ in this paper, this technique could also be used for the general case of complex reflection groups.

Given the symmetric group $\Sigma_n$ and a representation $\h$ of $\Sigma_n$ over a field $K$, we let $\mathcal{S}$ be the set of transpositions in $\Sigma_n$. For each $s \in \mathcal{S}$ we assign a vector $\alpha_s \in \h^*$ spanning the image of $1-s$, and choose $\alpha_s^\vee \in \h$ so that $(1-s)x=\langle \alpha_s^\vee,x\rangle \alpha_s$ for all $x \in \h^*$, where $\langle \cdot,\cdot\rangle$ indicates the pairing between $\h$ and $\h^*$.

Let $T(\h \oplus \h^*)$ be the tensor algebra of $\h \oplus \h^*$. We choose $\hbar,c \in K$. Then we define the {\it rational Cherednik algebra of type A $\HH_{\hbar,c}(\h)$} as the quotient of $K[\Sigma_n] \ltimes T(\h \oplus \h^*)$ by the following relations for all $x,x' \in \h^*, y,y' \in \h$:
\begin{align*}
[x,x']=0, \quad [y,y' = 0], \quad [y,x] = \hbar\langle y,x\rangle - \sum_{s \in \mathcal{S}} c\langle y,\alpha_s\rangle\langle \alpha_s^\vee,x\rangle s.
\end{align*}
We can give $\HH_{\hbar,c}(\h)$ a $\mathbb{Z}$-grading by setting $\deg x=1$ for $x \in \h^*$, $\deg y = -1$ for $y \in \h$, and $\deg g=0$ for $g \in K[G]$. We get the PBW-type decomposition $\HH_{\hbar,c}(\h)=\Sym(\h) \otimes_K K[\Sigma_n] \otimes_K \Sym(\h^*)$ (\cite{EM}, section 3.2).

In general, for any $\alpha \ne 0$, $\HH_{\hbar,c}(\h)\simeq \HH_{\alpha\hbar,\alpha c}(\h)$. Then we can assume $\hbar=0$ or $\hbar = 1$.

Let $\tau$ be a representation of $G$. The {\it Verma module} $M_{\hbar,c}(\h,\tau)$ is defined as $\HH_{\hbar,c}(\h) \otimes_{K[\Sigma_n] \ltimes \Sym(\h)} \tau$. Using the PBW decomposition of the Cherednik algebra, we see that $M_{\hbar,c}(\h,\tau)=\Sym(\h^*) \otimes_K \tau$ as a $K$-vector space; we can give this a $\mathbb{Z}$-grading in an obvious way. We are concerned with the case $\tau$ trivial, or the {\it polynomial representation} $\Sym(\h^*)$, which is a polynomial ring. We refer to the polynomial representation as $A$.

As described in section 2.5 of \cite{BC1}, $A=\Sym(\h^*)$ has a unique maximal graded proper submodule $J$ which can be realized as the kernel of the contravariant form $\beta_c: \Sym(\h^*) \times \Sym(\h) \to k$; $\beta_c$ can be characterized by the property that for all $x \in \h^*, y \in \h, f \in \Sym(\h^*), g \in  \Sym(\h)$:
\begin{align*}
\beta_c(xf,g)=\beta_c(f,xg), \quad \beta_c(f,yg) = \beta_c(yf,g), \quad \beta_c(1,1) = 1.
\end{align*}
The quotient $A/J$ is a finite-dimensional irreducible $\mathbb{Z}$-graded representation of $\HH_{\hbar,c}(\h)$. The {\it Hilbert series} $h_{A/J}(t)$ of $A/J$ is $\sum_{j=0}^\infty (\dim_k L_j)t^k$ where $L_j$ is the $j$-graded factor of $A/J$.

 We call $c$ {\it generic} if we do not specify its value. We will be concerned with the case $\hbar=1$, $c$ generic, and the characteristic $p$ of $K$ dividing $n$. We also have $\dim \h = n-1$, so $\h$ is an irreducible representation of the symmetric group. We state our result for this case below:

 \newtheorem*{thm:main}{Theorem \ref{thm:main}}
\begin{thm:main} The irreducible representation $A/J$  of $\HH_{\hbar,c}(\h)$ is a complete intersection; it has Hilbert series $\left(\frac{1-t^p}{1-t}\right)^{n-1}$. $J$ is generated by the coefficients of $z^p$ in the formal power series \\
$F_a=\frac{1}{1-x_az} \left( \sum_{k=0}^{p-1} \binom{c}{k}(-1+\prod_{j=0}^{n-1} (1-x_jz))^k\right)$ for $a=0,\dots,n-2$. 
\end{thm:main}

In this paper we consider the case where the characteristic of $K$ divides $n$. This is related to the case in characteristic $0$ where $c$ takes the specific value $p/n$, as described in \cite{CE}. In that case the generators of the ideal $J$ were the residues at infinity of $\frac{1}{z-x_a} \prod_{i=0}^{n-1} (z-x_i)^c$ for each $a$. To get a similar result in positive characteristic, we must consider formal power series in $z$ with coefficients from $A$. The formal power series would be $\frac{1}{z^{p+1}}\frac{1}{1-x_az}\prod_{j=0}^{n-1} (1-x_jz)^c$, with the corresponding generator as the coefficient of $1/z$. We simplify and truncate the formal power series so we can define it in positive characteristic.

\paragraph{Acknowledgements} This project was done with the support of the Undergraduate Research Opportunities Program (UROP) at MIT. The author thanks Pavel Etingof and Yi Sun for supervising this project. The computer algebra system SAGE was used for obtaining data.

 \subsection{Definitions}

To understand the action of $\HH_{\hbar,c}(\h)$ on $A$, we can use the PBW decompositions. The action of $\Sym(\h^*)$ on $A$ is by left multiplication; $K[\Sigma_n]$ acts by the diagonal action, and $\Sym(\h)$ acts via {\it Dunkl operators}. For $y \in \h$, the Dunkl operator $D_y$ acts on $A$ by:
\begin{align*}
D_y(f) = \hbar \partial_y f  - \sum_{s \in \mathcal{S}} c \frac{ ( y, \alpha_s )}{\alpha_s} (1-s). f.
\end{align*}

We choose bases for $\h,\h^*$ in the following way: let $V$ be the vector space spanned by $y_0,\dots,y_{n-1}$ and let $\h$ be the subspace spanned by $y_i-y_j$ for $i \ne j$; $\Sigma_n$ acts by permuting indices. Then if $x_0,\dots,x_{n-1}$ is the dual basis for $V^*$, we see that $\h^*$ is the span of $x_0,\dots,x_{n-1}$ under the relation $x_0+\dots+x_{n-1}=0$; alternatively we can consider $\h^*$ as the span of $x_0,\dots,x_{n-2}$ with $x_{n-1}$ defined as $-x_0-\dots-x_{n-2}$.  If $D_i$ is the Dunkl operator corresponding to $y_i$, the Dunkl operators for the elements of $\h$ are spanned by $D_i-D_j$ with $i \ne j$.

For a transposition $s_{ij} \in \Sigma_n$ with $i<j$, we let the corresponding vector $\alpha_{s_{ij}} \in \h^*$ be $x_i-x_j$.

The characteristic of the field $K$ is $p>0$. We assume $p>2$ since the case $p=2$ is fully characterized in \cite{L}.

@@

@@

Because of the definition of the contravariant form $\beta_c$, showing that an element $f$ of $A$ is in the kernel of $\beta_c$ is equivalent to showing that the Dunkl operators corresponding to the basis elements of $\h$ annihilate $f$.

@@

\begin{proposition}\label{prop:linind} The $f_a$ for $a=0,\dots,n-2$ are linearly independent homogeneous polynomials of degree $p$.
\end{proposition}

@@

@@

We write $x$ for the vector $\langle x_0,\dots, x_{n-2} \rangle$, where the $x_i$ are taken from the rational function field in $c$ over $K$. Then we can consider $f_a$ as a function on these vectors $x$ for all $a$. For any rational function $u( c)$, we let $u(c )x=\langle u(c )x_0,\dots, u(c )x_{n-2}\rangle$.

@@

Using these propositions, we are able to prove the main theorem.

\begin{theorem}\label{thm:main} The irreducible representation $A/J$  of $\HH_{\hbar,c}(\h)$ is a complete intersection; it has Hilbert series $\left(\frac{1-t^p}{1-t}\right)^{n-1}$. $J$ is generated by the coefficients of $z^p$ in the formal power series
\\$F_a=\frac{1}{1-x_az} \left( \sum_{k=0}^{p-1} \binom{c}{k}(-1+\prod_{j=0}^{n-1} (1-x_jz))^k\right)$ for $a=0,\dots,n-2$. 
\end{theorem}

\begin{proof} It suffices to show that the $f_a$ generate the ideal $J$ and that $A/J$ has Hilbert series $\left(\frac{1-t^p}{1-t}\right)^{n-1}$.

By Propositions~\ref{prop:linind}, \ref{prop:ci}, $A/I$ is a complete intersection with $n-1$ generators of degree $p$. It then must have Hilbert series $h_{A/I}(t)=\left(\frac{1-t^p}{1-t}\right)^{n-1}$. By Proposition~\ref{prop:ann}, the generators of $I$ are annihilated by the Dunkl operators, so $I \subseteq J$.

@@
